manuscraps

### 2025-07-04 ms/manuscraps.qmd
added aa_loo1 table to SI

---
title: Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves
format:
  pdf:
    keep-tex: true  
    toc: false
author:
  - name: "Christopher D. Muir$^{1,2*}$"
    orcid: 0000-0003-2555-3878
  - name: "Wei Shen Lim$^{1}$"
  - name: "Dachuan Wang$^{1,2}$"
affiliations: false
number-sections: false
header-includes: |
  \usepackage{hyperref}
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
  \newcommand{\aalow}{\mathrm{AA}_{150}}
  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
  \newcommand{\aax}{$\mathrm{AA}$}
  \newcommand{\aay}{\mathrm{AA}}
  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
  \newcommand{\aqcurve}{$A \textendash Q$}
  \newcommand{\ca}{$C_\mathrm{a}$}
  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
  \newcommand{\gsw}{$g_\text{sw}$}
  \newcommand{\loggsw}{$\log(g_\text{sw})$}
  \newcommand{\logA}{$\log(A)$}
  \newcommand{\ppfd}{$\mathrm{PPFD}$}
  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
  \newcommand{\rh}{$\mathrm{RH}$}
  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
mainfont: Times New Roman
bibliography: solanum-aa.bib
csl: https://www.zotero.org/styles/science
---

```{=latex}
\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}
```

```{r load-objects, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(brms)
library(cowplot)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(knitr)
library(lubridate)
library(magrittr)
library(readr)
library(stringr)
library(tibble)
library(tidyr)

theme_set(theme_cowplot())

source("../r/functions.R")

aa_args = read_rds("../objects/aa_args1.rds")
aa_stats = read_rds("../objects/aa_stats.rds")
# sort list by name
aa_stats = aa_stats[order(names(aa_stats))]

plant_info = read_rds("../data/plant-info.rds")

accession_info = read_rds("../data/accession-info.rds") |>
  dplyr::select(species, accession)

rh_curves = read_rds("../data/trimmed_rh_curves.rds")

stomata = read_rds("../data/stomata.rds")

fit_aa1 = read_rds("../objects/fit_aa1.rds")
fit_aa2 = read_rds("../objects/fit_aa2.rds")
dim_fit_aa = dim(fit_aa1$fit)

df_aa_pred1 = read_rds("../objects/df_aa_pred1.rds")
# df_aa_pred2 = read_rds("../objects/df_aa_pred2.rds")
df_aa_pred3 = read_rds("../objects/df_aa_pred3.rds")

n_plant = length(unique(rh_curves$acc_id))
n_acc = length(unique(rh_curves$acc))
n_sp = accession_info |>
  filter(accession %in% unique(rh_curves$acc)) |>
  pull(species) |>
  unique() |>
  length()

# average number of individual replicates for accession per light treatment
mean_n_rep = rh_curves |>
  summarize(
    n = length(unique(acc_id)),
    .by = c("acc", "light_treatment")
  ) |>
  pull(n) |>
  mean() |>
  prettyNum(digits = 3)

licor_date_range = plant_info |>
  filter(acc_id %in% unique(rh_curves$acc_id)) |>
  dplyr::select(`1s_rh_response_date`, `2s_rh_response_date`) |>
  pivot_longer(cols = everything()) |>
  pull(value) |>
  ymd() |>
  range(na.rm = TRUE)

n_weeks = prettyNum(as.numeric(difftime(licor_date_range[2], licor_date_range[1], units = "days")) / 7, digits = 3)
 # number of weeks for LICOR measurements
start_licor_date = format(licor_date_range[1], "%B %d, %Y")
end_licor_date = format(licor_date_range[2], "%B %d, %Y")

# Environmental data
df_germ_summary = read_rds("../data/df_germ_summary.rds") |>
  ungroup() |>
  summarize(value = median(value), .by = c("period", "variable", "parameter"))

df_growth_summary = read_rds("../data/df_growth_summary.rds") |>
  ungroup() |>
  filter(!is.na(value))

# environmental condition in germination room

germ_temp_day = df_germ_summary |>
  filter(period == "day", variable == "degC", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

germ_temp_night = df_germ_summary |>
  filter(period == "night", variable == "degC", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

germ_rh_day = df_germ_summary |>
  filter(period == "day", variable == "RH", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

germ_rh_night = df_germ_summary |>
  filter(period == "night", variable == "RH", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

germ_ppfd = df_germ_summary |>
  filter(period == "day", variable == "PAR", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

# environmental conditions in growth room

d1 = df_growth_summary |>
  summarize(value = median(value),
            .by = c("period", "rack", "variable", "parameter"))

ppfd_sun = d1 |>
  filter(period == "day", rack == "lower", variable == "PAR", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

ppfd_shade = d1 |>
  filter(period == "day", rack == "upper", variable == "PAR", parameter == "mu") |>
  pull(value) |>
  prettyNum(digits = 3)

# degrees warmer that the sun treatment air was
temp_diff = df_growth_summary |>
  filter(period == "day", rack != "hobo", variable == "degC", parameter == "mu") |>
  pivot_wider(names_from = "rack") |>
  mutate(diff = lower - upper) |>
  summarize(value = median(diff, na.rm = TRUE)) |>
  pull(value) |>
  prettyNum(digits = 3)

# difference in RH
rh_diff = df_growth_summary |>
  filter(period == "day", rack != "hobo", variable == "RH", parameter == "mu") |>
  pivot_wider(names_from = "rack") |>
  mutate(diff = upper - lower) |>
  summarize(value = median(diff, na.rm = TRUE)) |>
  pull(value) |>
  prettyNum(digits = 3)

df_leaf_temp = plant_info |>
  filter(!is.na(leaf_temperature_C)) |>
  summarize(
    leaf_temperature_C = mean(leaf_temperature_C),
    n = n(),
    .by = "light_treatment"
  )

# difference in leaf temperature between treatments
leaf_temp_diff = df_leaf_temp$leaf_temperature_C |>
  diff() |>
  prettyNum(digits = 3)

# number of leaves for which we have measured leaf temp under ambient conditions
n_leaf_temp = sum(df_leaf_temp$n)

# data on leaflet type
df_leaflet_type = plant_info |>
  summarize(n = n(), .by = "leaflet") |>
  pivot_wider(names_from = "leaflet", values_from = "n") |>
  mutate(
    percent_terminal = prettyNum(terminal / (terminal + lateral) * 100, digits = 3),
    percent_lateral = prettyNum(lateral / (terminal + lateral) * 100, digits = 3)
  )

percent_terminal = df_leaflet_type$percent_terminal
percent_lateral = df_leaflet_type$percent_lateral
# percent of time that anatomical traits were measured on different leaflet than LICOR measurements
percent_change_leaflet = full_join(
  plant_info |>
    filter(acc_id %in% unique(rh_curves$acc_id)) |>
    select(acc_id, licor = leaflet),
  stomata |>
    mutate(acc_id = str_c(accession, replicate, sep = "-")) |>
    filter(acc_id %in% unique(rh_curves$acc_id)) |>
    select(acc_id, stomata = leaflet),
  by = join_by(acc_id)
) |>
  summarize(x = mean(licor != stomata, na.rm = TRUE)) |>
  pull(x) |>
  multiply_by(100) |>
  prettyNum(digits = 3)

 # percent of times where AA and AQ curves were done on same leaflet
percent_aa_and_aq = plant_info |>
  filter(acc_id %in% unique(rh_curves$acc_id)) |>
  mutate(same_leaflets = !switch_leaflets) |>
  pull(same_leaflets) |>
  mean() |>
  multiply_by(100) |>
  prettyNum(digits = 3)

# data on chamber light conditions
percent_red = rh_curves |>
  mutate(p_red = Q_red / Q) |>
  summarize(p_red = mean(p_red, na.rm = TRUE),
            .by = c(acc_id, light_intensity)) |>
  summarize(p_red = mean(p_red, na.rm = TRUE), .by = light_intensity) |>
  pivot_wider(names_from = light_intensity, values_from = p_red)

p_red_high = format(signif(100 * percent_red$`2000`, 3), nsmall = 1)
p_blue_high = format(signif(100 * (1 - percent_red$`2000`), 3), nsmall = 1)
p_red_low = format(signif(100 * percent_red$`150`, 3), nsmall = 1)
p_blue_low = format(signif(100 * (1 - percent_red$`150`), 3), nsmall = 1)

# percent measure LMA on alternate leaflet
percent_lma_switch = plant_info |>
  filter(acc_id %in% rh_curves$acc_id) |> 
  summarize(x = mean(leaflet != lma_leaflet, na.rm = TRUE)) |>
  pull(x) |>
  multiply_by(100) |>
  prettyNum(digits = 3)

 # percent leaves missing LMA
percent_lma_missing = plant_info |>
  filter(acc_id %in% rh_curves$acc_id) |> 
  summarize(x = mean(is.na(lma_leaflet))) |>
  pull(x) |>
  multiply_by(100) |>
  prettyNum(digits = 3)

### Cleaning RH curves
ca_min = aa_args$ca_min
ca_max = aa_args$ca_max
n_min_points_per_rh_curve = aa_args$n_min_points_per_rh_curve
point_outlier_threshold = aa_args$point_outlier_threshold
aa_outlier_threshold = aa_args$aa_outlier_threshold
thinning_interval = aa_args$thinning_interval

### Cleaning AQ curves
aq_r2_threshold = aa_args$aq_r2_threshold

### Bayesian data analysis
n_model = 5 |> # i don't think I need this anymore
  number_to_word()

### Summarizing AA results
aa_mean_range = df_aa_pred1 |>
  ungroup() |>
  mutate(signif = sign(.lower) == sign(.upper)) |>
  summarize(
    min = min(aa),
    max = max(aa),
    aa = mean(aa),
    n_signif = sum(signif),
    n_acc = n(),
    .by = c(light_treatment, light_intensity)
  ) |>
  mutate(x = sprintf(
    "%.3f (range: %.3f–%.3f; %.0f of %.0f populations significant)",
    aa,
    min,
    max,
    n_signif,
    n_acc
  ))

aa_summary3 = df_aa_pred3 |>
  mutate(signif = (.upper < 0)) |>
  summarize(n_signif = sum(signif), .by = light_treatment) |>
  pull(n_signif, light_treatment)

# Effect of light treatment on LMA
post_lma = posterior_summary(fit_aa2)["b_llma_light_treatmenthigh", ]

## Abstract

The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts. Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species. This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand. We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides*). Plant grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity or native light habitat. Developmental plasticity in leaf thickness and/or density, an adaptive response to optimize light interception, likely affects CO$_2$ diffusion within the leaf, altering amphistomy advantage as a byproduct. Developmental plasticity may therefore play an important and previously unnoticed role in explaining the adaptive significance of amphistomy and the distribution of amphistomatous leaves in nature.

## Main text

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022]. Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], and improving crops [@hofmann_impact_2025]. Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].

<!-- Paragraph 2 -->
<!-- Parkhurst gradients 1988 -->
<!-- Syvertsen et al 1995 - may be more here to check about gi in sun and shade-->
<!-- "Farquhar and Raschke (2). However, using a technique similar to ours, Sharkey et al. (11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020]. Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_environmental_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface. All else being equal, gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. The extent to which amphistomy increases CO$_2$ supply depends on resistance to diffusion in intercellular airspaces. This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@parkhurst_intercellular_1990]. Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_is_2019].

<!-- Paragraph 3 -->
The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019]. But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025]. Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface [@lyshede_comparative_2002], suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous [@salisbury_i_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018]. Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

<!-- Paragraph 4+, hypotheses and predictions -->
The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986]. We quantify this benefit as the amphistomy advantage (\aax{}), the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf [@parkhurst_adaptive_1978; @triplett_amphistomy_2024]. Why would \aax{} be greater in sun than shade? We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive' ([Fig. @fig-concept]).

*Acclimatory hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand). A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase \aax{}, all else being equal (Supporting Information). If the acclimatory hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.

*Plasticity hypothesis*: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019]. Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014]. Plastic changes in leaf anatomy and biochemistry could modulate \aax{} as a byproduct. Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand. Under the plastic hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities. Secondarily, $\aasun{}$ and \gmaxsun{} should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.

*Constitutive hypothesis*: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes. We therefore predict genotypes from more sunny, open habitats will have consistently greater \aax under also measurement and growth light intensities. For herbaceous plants, light intensity is largely a function of the tree canopy [@engelbrecht_evaluation_2001]. Herbs growing in the open will regularly experience high light intensity; herbs growing under a forest canopy will often experience low light intensity.

```{r, concept, echo = FALSE}
#| label: fig-concept
#| fig-align: center
#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatory hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI). The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI. The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."

knitr::include_graphics("../figures/amphisomy draft 2024-03-28.jpg")
```

The primary directional predictions for each hypothesis are summarized in @tbl-hypotheses; detailed predictions for results that would indicate support for multiple hypotheses are in @tbl-predictions.

```{r, hypotheses, echo = FALSE}
#| label: tbl-hypotheses
#| tbl-cap: "Three nonmutually exclusive hypotheses and directional predictions explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats. For each hypothesis, we make predictions for how measurement light intensity (\\ppfdequals{150} vs. \\ppfdqty{2000}), growth light intensity (sun vs. shade), and native plant area index (PAI) would affect \\aax{}. \\ppfd{}: photosynthetic photon flux density."

kableExtra::kable(
  data.frame(
    hypothesis = c("acclimatory", "plastic", "constitutive"),
    measurement = c(
      "$\\mathrm{AA}_{2000} > \\mathrm{AA}_{150}$",
      "$\\mathrm{AA}_{2000} = \\mathrm{AA}_{150}$",
      "$\\mathrm{AA}_{2000} = \\mathrm{AA}_{150}$"
    ),
    growth = c(
      "$\\mathrm{AA}_{\\text{sun}} = \\mathrm{AA}_{\\text{shade}}$",
      "$\\mathrm{AA}_{\\text{sun}} > \\mathrm{AA}_{\\text{shade}}$",
      "$\\mathrm{AA}_{\\text{sun}} = \\mathrm{AA}_{\\text{shade}}$"
    ),
    native = c(
      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) = 0$",
      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) = 0$",
      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) > 0$"
    )
  ),
  col.names = c(
    "Hypothesis",
    "Measurement light intensity",
    "Growth light intensity",
    "Native PAI"
  ),
  escape = FALSE,
  booktabs = TRUE,
  format = "latex",
  align = "c"
) |>
  kable_styling() |>
  column_spec(1, width = "1in") |>
  column_spec(2, width = "1.5in") |>
  column_spec(3, width = "1.5in") |>
  column_spec(4, width = "1.25in")
```

<!-- Paragraph 5 (summarize methods) -->
We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (). We measured AA on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]. With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details). We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods). The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center. The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see materials and methods in supplementary materials). The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.

<!-- Paragraph 6 (summarize main results) -->
Consistent with biophysical theory of CO$_2$ diffusion within leaves, $\aay{} > 0$ for all populations ([Fig. @fig-aa]A). \aax{} varied substantially between measurement light intensities, growth light intensities, and among populations (point to model comparison evidence). Measured under high light intensity, \aax{} was consistently greater for sun plants. The average \aax{} among populations in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same populations grown at high light intensity showed a mean \aax{} of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`. Contrary to the predictions of the assimilatory hypothesis, \aax{} was greater in all populations under low measurement light intensity for both sun and shade grown plants. The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to to `r aa_summary3["low"]` populations for shade-grown plants. The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively. There was a slight tendency for populations from more closed habitats (greater PAI), but \aax{} varied widely among populations from open habitats (low PAI) regardless of growth and measurement light intensities ([Fig. @fig-aa]B).

```{r, aa, echo = FALSE, warning=FALSE, message=FALSE} 
#| label: fig-aa
#| fig-align: center
#| fig-width: 6
#| fig-height: 8
#| fig-cap: "Caption on following page."

plot_grid(
  read_rds("../objects/fig_aa.rds"),
  read_rds("../objects/fig_aa_pai.rds"),
  ncol = 1,
  labels = "AUTO",
  label_size = 20,
  label_fontface = "bold",
  align = "hv"
)

```{=latex}
\addtocounter{figure}{-1}
```

```{r, aa_cap, echo = FALSE, warning=FALSE, message=FALSE} 
#| label: fig-aacap
#| fig-align: center
#| fig-width: 0.1
#| fig-height: 0.1
#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in wild tomatoes is amplified under simulated sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf. In both panels, estimates are shown for plants grown under simulated shade or sun and measured under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}). The dashed line at zero indicates no difference in photosynthesis between amphi- and pseudohypostomatous leaves. (\\textbf{A}) The points are the posterior median \\aax{} for each population, with error bars showing 95% confidence intervals. Within each facet, the populations are arranged by \\aax{} estimated in that growth and measurement condition. (\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale). The confidence intervals were omitted for visual clarity."

ggplot()
```

<!-- Paragraph 7 (summarize secondary results) -->
The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis. Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on \aax{}. LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019]. When LMA is included in the model, the effect of growth light intensity on \aax{} is no longer significant based on model comparison (Table S8) and LMA is weakly positively correlated with \aax{} ([Fig. @fig-lma_aa]). In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table S8). Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}. The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972]. At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted. Hence, amphistomy may, expectedly, be particularly beneficial for sun leaves experiencing intermittant shade or cloud cover. Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface. Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.

```{r, lma_aa, echo = FALSE,warning=FALSE, message=FALSE} 
#| label: fig-lma_aa
#| fig-align: center
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated population-level LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}). The points are colored by growth light treatment. The solid line is the posterior median of the linear regression of individual-level LMA on $\\aahigh$  with the shaded region showing the 95% confidence ribbon; the dashed line is the same for $\\aalow$. Confidence intervals for each population estimate and individual-level estimates are omitted for visual clarity."

read_rds("../objects/fig_lma_aa.rds")

We conclude that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives. Our results advance existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather that anatomical and biochemical change associated with higher light intensity modulate \aax{}. Developmental plasticity needs to be considered when interpreting phylogenetic comparative analyses of adaptation, because plastic differences may be mistaken for genetic differences.

maybe this wording is better: Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites). If our estimates of AA from the wild tomato clades

A second major conclusion is that the significant photosynthetic gain of amphistomatous leaves implies that the many hypostomatous leaves that dominate mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss. Under realistic assumptions, we estimate that hypostomatous leaves could achieve the same $A$ with X to X% less water loss if they were amphistomatous (Table S9). Ours is the first experimental demonstration that amphistomy is beneficial for sun leaves because of plastic changes, but future research needs to estimates the tradeoffs associated with amphistomatous leaves in order to fully explain their rarity among many woody plants and shade-tolerant herbs.

## Acknowledgements

Sam McKlin and Tom Buckley helped with protocol development. Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai Yasuda contributed to data collection. Sarah Friedrich helped with conceptual figure.

### Funding:

US National Science Foundation OIA-1929167 to C.D.M.

### Author contributions:

Conceptualization: C.D.M.; Methodology: C.D.M., W.S.L.; Investigation: C.D.M., W.S.L., D.W.; Visualization: C.D.M.; Funding acquisition: C.D.M.; Writing – original draft: C.D.M.; Writing – review & editing: C.D.M., W.S.L., D.W.

## References

::: {#refs}
:::

{{< pagebreak >}}

# Supplementary Materials

```{=latex}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}
```

## Materials and Methods {#sec-methods}

### Populations

We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect. *Lycopersicon* and sect. *Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S. lycopersicum* var. *lycopersicum* (@tbl-populations). Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`. Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession. [anything else to say here? maybe explain population selection and phylogeny?]

```{r, populations, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-populations
#| tbl-cap: "Accession information of *Solanum* populations used in this study. The species name, accession number, collection latitude, longitude, elevation, and daily solar radiation estimated using the SPLASH algorithm (see 'Climate data'). TGRC: Tomato Genetics Resource Center; PAI: Plant area index."

# Copied from solanum-aq project

accession_info = read_rds("../data/accession-info.rds") |>
  filter(accession %in% rh_curves$acc) |>
  left_join(read_rds("../data/accession-gedi.rds") |>
              select(accession, pai), by = join_by(accession)) |>
  arrange(species, accession) |>
  select(
    Species = species,
    `TGRC accession` = accession,
    `Latitude` = lat,
    `Longitude` = lon,
    `Elevation (mas)` = elev,
    `$\\mathrm{PAI}~(\\unit{\\meter\\squared\\per\\meter\\squared})$` = pai
  )

>>>>

---
title: manuscraps
format:
  pdf:
    keep-tex: false  
    toc: false
number-sections: false
bibliography: solanum-aa.bib
csl: https://www.zotero.org/styles/science
---
